Up to 25\% of beneficial mutations in protein sequences are not adaptive innovations in mammals

### 2023-02-01 manuscript/back-mutations-intro.tex
Introduction with focus on back-mutations as a separate file

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}

\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
\addbibresource{codon_models.bib}

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}

\newcommand{\NS}[1]{\textcolor{red}{\textbf{\emph{[NS: #1]}}}}

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\textrm{Var}}
\newcommand{\der}{\textrm{d}}
\newcommand{\e}{\textrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\textrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\textrm{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\renewcommand{\baselinestretch}{1.5}
\linenumbers

\title{Up to 25\% of beneficial mutations in human protein sequences are not adaptive innovations in mammals}

\author{
    \large
    \textbf{T. {Latrille}$^{1}$, J. {Joseph}$^{2}$, D.~A. {Hartasánchez}$^{1}$, N. {Salamin}$^{1}$}\\
    \normalsize
    $^{1}$Université de Lausanne, Lausanne, Switzerland\\
    $^{2}$Université de Lyon, CNRS, LBBE UMR 5558, Villeurbanne, France \\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

% The submission checklist is available at:
% https://docs.google.com/document/d/1aHCF1on0mHTK2DoroPCLP_xto4xwHnGJCGwaL1Vm2UE/edit?usp=sharing

\begin{document}
    \maketitle

%TC:endignore
    \begin{abstract}
        Mutations can be beneficial by bringing an innovation to their bearer, allowing them to adapt to changes in environments or in selective pressure. However, mutations can also be beneficial because they are repairing previous deleterious changes, simply restoring existing functions. In this study, we first estimated selective effects of mutations inside mammalian protein coding sequences, under a model assuming no adaptation at the phylogenetic scale. We then estimated the proportion of beneficial mutations that are not adaptive innovations among all beneficial mutations at the population scale. Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated. In modern humans, it results in around 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.
    \end{abstract}

@@

    Adaptation is considered one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
    It enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859} [NS: maybe cite Losos or some key papers on ecological speciation rather than Darwin twice].
    In evolution, adaptation is tightly linked to the notion of change at the level of the environment and to species responding to this change~\cite{merrell_adaptive_1994}.
    The process of adaptation leaves marks of accelerated evolution in genomes since mutations that are beneficial under new conditions fix faster than neutral mutations (fig.~\ref{fig:fitness-landscape}A).
    The signature of an accelerated rate of substitutions, defined as the rate of mutation times the probability of fixation, can thus be interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
    The current availability of large-scale genomic data and the development of theoretical models have allowed to detect and quantify adaptation across genes and lineages, and these analyses have become common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019} [NS: cite the paper by Anna Marcionetti here].
    Although these approaches have led to a better understanding of the processes determining the rates of molecular evolution, they have had the effect of associating beneficial mutations with adaptive evolution, while disconnecting them from the notion of change defining adaptation.
    We argue that we should limit the use of adaptive mutations to those that are associated with adaptation as such and dissociate beneficial mutations from adaptive mutations since adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    We propose here a new approach to identify the mechanisms underlying the origin of beneficial mutations.
    In particular, we suggest that many beneficial mutations are not adaptive, but are rather restoring ancestral states of higher fitness.
    We used large-scale genomic data to integrate analyses at both the phylogenetic and population scales to differentiate between truly adaptive mutations from mutations restoring the ancestral fitness.
    Our results show that up to a quarter of beneficial mutations across mammal species are not driven by adaptive innovations, but rather compensate the effect of ancestral mutations to recover the optimal state of the species.

    In a constant environment, if a population is already adapted to these conditions, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a new mutation that is restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    The restoration of the ancestral fitness can happen at a different locus than the initial mutation, in which case it is referred to as a compensating [NS: ou compensatory ?] mutation~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    However, if the restoration of the ancestral fitness happens at the same locus than the initial mutation, in which case it is referred to as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}.
    In this study, we focus on the latter case.
    While compensatory mutations induce genetic changes at the sequence level, and thus genetic diversification, beneficial back-mutations reduce the genetic diversity and do not contribute to genetic innovation.
    They merely cancel the effect of the previous deleterious changes.
    As a consequence, not all beneficial mutations contribute to genetic innovation and truly adaptive evolution can only be correctly estimated if we can account for the number of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    % Philosophically, our distinction between beneficial back-mutations and adaptation can appear to be far-fetched, and one can argue that beneficial back-mutations is adaptation to its own deteriorating genome.
    % However, there is a difference between adapting to a mutation, for instance by means of a compensatory mutation elsewhere, and cancelling the mutation, by means of a substitution on the misfit amino-acid.

    %For an individual carrying a beneficial mutation, whether this mutation is a new innovation responding to a change in environment or a beneficial back-mutation is indistinguishable~\cite{charlesworth_other_2007}.
    Given a beneficial mutation, distinguishing whether it is a new innovation responding to a change in environment or a beneficial back-mutation is impossible with only one individual at hand~\cite{charlesworth_other_2007}.
    This limitation persists at the level of the population, since both types of mutation will result in a positive transmission bias of the beneficial allele.
    However, at the macro-evolutionary scale the consequences of these two types of mutations are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and preserve well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and, hence, in the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known because we will expect to see changes from non-optimal amino acids to optimal ones.
    If we wish to quantify the amount of beneficial back-mutations that have happened along a phylogeny we need to identify the selection coefficients of mutations and to achieve this we require the fitness landscape to be stable and be known.

    The placental mammals represent an excellent study system to perform this analysis.
    Having originated ~100 Mya\cite{kumar_timetree_2017}, they diversified quickly.
    Additionally, we have polymorphism data available for many species\cite{howe_ensembl_2021} and high quality protein-coding DNA alignments across the genome\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
    By restricting our analysis to protein-coding orthologous genes to all the phylogeny, and further excluding any genes for which there have been claims of pervasive directional selection, we can not only assume a constant fitness landscape for each of them throughout placental mammal evolution, but actually reconstruct the fitness landscape from variable positions within the sequence along the phylogeny.
    For each gene, we used a codon model of mutation-selection~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014} to obtain relative fitnesses for all amino-acids for each site of the sequence by fitting the model to the multi-species sequence alignment (fig~\ref{fig:method}A) and assuming that the underlying fitness landscape is stable along the phylogenetic tree.
    %Mutation-selection codon models provide a nearly neutral framework of evolution taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.
    From the fitness landscape, we can calculate the scaled selection coefficient $\Sphy$ for any possible amino-acid change at each position.
    Under this model, a mutation with a positive $\Sphy$ would be driving the current site toward a known fitter amino-acid and would thus be considered as a beneficial back-mutation (fig~\ref{fig:method}B), given the assumed stability of the fitness landscape.

@@

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site of a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
            At the genome level, slightly deleterious mutations that have reached fixation are supposedly scattered at different loci, such that beneficial back-mutations should also be spread across the genome.
            We thus expect that the genome-wide signature of beneficial back-mutations can be detected and quantified, even though each mutation has individually a small beneficial effect on their bearer.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba}
        \caption{
            Integrating datasets at different scales.
            At the phylogenetic level (panel A), the amino acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            Mutation-selection models assume a stable fitness landscape, taking into account only non-adaptive evolution.
            A mutation with a $\Sphy > 1$ would be driving the current site toward are fitter amino-acid, and such mutation are considered a beneficial back-mutation repairing existing functions (panel B).
            At the population scale (panel C), each single nucleotide polymorphism (SNP) segregating in the population can be classified as a beneficial back-mutation (in blue) or not.
            The number of derived alleles allows us to estimate the probability of beneficial mutations at the population scale, for all SNPs or among back-mutations (panel D).
            The probability of back-mutation is obtained as their fraction among all possible mutations from the ancestral genome sequence (panel E).
            Altogether, we can estimate the probability of back-mutations among all beneficial ones, using Bayes' theorem (panel F).
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{Back-mutations in the terminal lineages}
    We first estimated the number of beneficial back-mutation in the terminal branches of mammals by reconstructing the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the phylogenetic tree of mammals (fig.~\ref{fig:method}A).
    Among all the substitutions that were found in the terminal branches, between 10 and 13\% had $\Sphy > 1$ (table~S1), while between 25 and 30\% had $\Sphy > 0$ (table~S2).
    For substitutions with $\Sphy > 1$, the $\dnds$ ratio of non-synonymous over synonymous divergence was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    Beneficial back-mutations with $\Sphy > 1$ therefore reached fixation more frequently than synonymous mutations that are supposed to be neutral, which means that these back-mutations are effectively beneficial.
    This result further indicated that $\dnds$ used to detect adaptation while relying on substitutions between closely related species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is biased due to the presence of beneficial back-mutations that are not adaptive.
    We can alleviate this bias by computing $\dnds$ exclusively on non-synonymous sites with a $\Sphy < 0$, which represents the rate of evolution when beneficial back-mutations are discarded.
    By comparing these two ways of estimating $\dnds$ (see Material \& Methods), we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations while they only represent between 1\% and 2\% of mutational opportunities (table~S2).

    \subsection*{Back-mutations in the populations}
    We then estimated the proportion of new mutations that were beneficial back-mutations while being under selection at the population scale (fig.~\ref{fig:method}C-E).
    We first reconstructed the ancestral genome (instead of the reference genome) of each mammal species with available polymorphism data.
    We computed the scaled selection coefficient ($\Sphy$) along this ancestral genome for each possible non-synonymous mutations, weighted by their mutational opportunity (see Material \& Methods).
    It allowed us to obtain the distribution of fitness effects (DFE) of new mutations from the ancestral sequence under the assumption of the stable fitness landscape as shown for \textit{Homo sapiens} in Fig.~\ref{fig:homo-afr-results}A.
    Because $\Sphy$ was based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ were considered, as shown above, beneficial back-mutations and only represented ca.~1.5\% of all possible new mutations (blue in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ca.~8.5\% of all observed mutations (blue in fig.~\ref{fig:homo-afr-results}B, table~S3).

    Within populations, the frequencies at which SNPs are segregating is also providing information on their selective effect.
    For instance, deleterious SNPs are segregating at lower frequencies because of the effect of purifying selection, which removes them from the population.
    It is possible to estimate the DFE at the population scale by gathering information across many SNPs~\cite{eyre-walker_distribution_2006, eyre-walker_estimating_2009, galtier_adaptive_2016, tataru_inference_2017} and these approaches offer a unique opportunity to compare selection coefficients estimated at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    We estimated the DFE at the population level and gathered robust statistical estimates by grouping SNPs observed in 5 classes of predicted selection coefficient (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}): severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
	%in mat/met
	%To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE~\cite{tataru_inference_2017, tataru_polydfe_2020}.
    %This model uses the count of derived alleles for each frequency class (fig.~\ref{fig:homo-afr-results}C) to infer the distribution of fitness effects (DFE).
    %Moreover, the model relies on the mutational opportunities ($L$) for each category of $\Sphy$ (fig.~\ref{fig:homo-afr-results}A, see Material \& Methods) as well as the count of synonymous derived alleles used as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral states.
    From the DFE, we further estimated the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogenetic tree of mammals were effectively purged at the population scale (red and yellow columns in fig.~\ref{fig:homo-afr-results}D)
    Purifying selection was therefore largely predictable and amino acids with low fitness across mammals were effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the phylogenetic scale were indeed beneficial for individuals currently bearing them (blue column in fig.~\ref{fig:homo-afr-results}D).
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange columns in fig.~\ref{fig:homo-afr-results}D) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (red and yellow rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (cyan and orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (blue row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity was associated with a smaller proportion of nearly-neutral mutations (fig.~\ref{fig:diversity}A\&B).
    This result confirmed the nearly-neutral theory and suggested that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) were more likely to discriminate between beneficial and deleterious mutations (fig~S1-S9), while mutations in populations with low diversity (e.g.~\textit{Homo}) were effectively nearly-neutral.

    \subsection*{Proportion of back-mutations among all beneficial mutations}

    Finally, we proposed an estimator for the probability of a beneficial mutation to be a beneficial back-mutation, computed as $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ (fig.~\ref{fig:method}C-F).
    We obtained from the analyses above the probabilities $\mathbb{P} [ \Sphy > 1 ]$, $\mathbb{P} [ \Spop > 1 ]$, $\mathbb{P} [ \Spop > 1  \given  \Sphy > 1]$.
    First, $\mathbb{P} [ \Sphy > 1 ]$ is the probability for a beneficial back-mutation ($\Sphy > 1$) from the ancestral sequence (blue fill of histogram in fig~\ref{fig:homo-afr-results}A and table~S3).
    Second, $\mathbb{P} [ \Spop > 1 ]$ is the probability of mutation to be beneficial at the level of the population ($\Spop > 1$), estimated from all available SNPs (black fill in fig~\ref{fig:diversity}A and table~S3).
    Finally, $\mathbb{P} [ \Spop > 1  \given  \Sphy > 1]$ is the probability for a back-mutation ($\Sphy > 1$) to be effectively beneficial at the level of the population ($\Spop > 1$), when we restricted our analysis on SNPs that were already predicted to be beneficial back-mutations (black fill for the category $\Sphy > 1$ in fig~\ref{fig:homo-afr-results}D and table~S3).
    Combining these probabilities using Bayes theorem allowed us to obtain the probability for a beneficial mutation to be a back-mutation $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ (see Material \& Methods).
    We estimated $\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ across all populations sampled and obtained a proportion that varied between 10 and 26\% (table~S3).

    The proportion of back-mutations among all beneficial ones did not show a clear relationship with the synonymous diversity within a population (Watterson $\theta_w$) (fig.~\ref{fig:diversity}C).
    Indeed, the relationship between $\Ne$ and the proportion of beneficial back-mutations is not trivial and should highly depend on time scales~\cite{charlesworth_other_2007}.
    A population that experienced a high $\Ne$ along its evolutionary time should be closer to the optimal state under a stable fitness landscape because of the higher efficacy of selection.
    This should decrease the probability of beneficial back-mutations~\cite{huber_determining_2017}.
    On the other hand, a high $\Ne$ over a short evolutionary time will induce more extreme fitness effects and a larger proportion of mutations will be beneficial (otherwise effectively neutral), thus increasing the proportion of beneficial back-mutations.
    Overall, the proportion of beneficial back-mutations is expected to be more dependent on long-term effective population size as well as contraction or expansion rather than on the short term effective population size captured by the synonymous diversity~\cite{charlesworth_other_2007,huber_determining_2017}.

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.homo-afr-results}
        \caption{
            Panel A: Distribution of fitness effects ($\Sphy$), predicted for all possible mutations away from the ancestral human genome.
            Mutations are divided into 5 classes of selection: severely deleterious (red), deleterious (yellow), weakly deleterious (light green), weakly beneficial (green) and beneficial (blue, supposedly beneficial back-mutations).
            Panel B. Distribution of fitness effects ($\Sphy$) for all observed SNPs in a sample of 8 individuals (out of 512 in the original dataset) of African descent.
            If they are less mutations observed than expected, this class is thus undergoing purifying selection.
            Panel B: The site-frequency spectrum (SFS) represents the proportion of mutations (y-axis) with a given number of derived alleles in the population (x-axis).
            SFS are drawn for a random sample of 16 alleles (mean in solid line and standard deviation in filled color) for each class of selection and for synonymous mutations which are supposedly neutral (black).
            At high frequencies, supposedly severally deleterious mutations are underrepresented.
            Panel D. For each class of selection (and for the set of all non-synonymous mutations), information from the SFS and the distribution of fitness effects are combined at the population scale to estimate the proportion of beneficial mutations $\PpolyAdv$, of nearly-neutral mutations $\PpolyNeutral$ and of deleterious mutations $\PpolyDel$.
        }
        \label{fig:homo-afr-results}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.heatmap}
        \caption{
            Reproducibility of results shown in figure~\ref{fig:homo-afr-results} in 28 populations across 6 genera.
            For each class of selection at the phylogenetic scale ($\Sphyclass \in \{ \divStrongDel, \divDel,  \divWeakDel,  \divWeakAdv, \divAdv \}$, in rows), site-frequency spectra and total number of sites are combined at the population scale to estimate the proportion of beneficial mutations ($\PpolyAdv$) in the top panel, of nearly-neutral mutations ($\PpolyNeutral$) in the middle panel and of deleterious mutations ($\PpolyDel$) in the bottom panel.
        }.
        \label{fig:heatmap}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.diversity}
        \caption{
            Panel A. For each population, the proportion of beneficial ($\PpolyAdv$), nearly-neutral ($\PpolyNeutral$) and of deleterious ($\PpolyDel$) mutations estimated at the population scale using all SNPs available.
            Population are sorted for increased synonymous diversity.
            Panel B. Proportion of nearly-neutral mutations ($\mathbb{P} [ -1 < \Spop < 1]$ in y-axis), shown as a function of the synonymous diversity (Watterson $\theta_w$ in x-axis).
            Panel C. Proportion of beneficial back-mutations among all beneficial mutations ($\mathbb{P} [ \Sphy > 1  \given  \Spop > 1]$ in y-axis), shown as a function of the synonymous diversity (Watterson $\theta_w$ in x-axis).
        }
        \label{fig:diversity}
    \end{figure*}

    \subsection*{A bridge between phylogenetic and population-genetics}
    In this study, we first estimated selective effects of mutations inside protein coding sequence, under a model assuming no adaptation at the phylogenetic scale.
    We then estimated the proportion of beneficial back-mutations, that are not adaptive innovations, and subsequently estimated their proportion among all beneficial mutations at the population scale.
    Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated, resulting in up to 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.

>>>>

\begin{abstract}
    Mutations can be beneficial by bringing an innovation to their bearer, allowing them to adapt to changes in environments or in selective pressure.
    However, mutations can also be beneficial because they are repairing previous deleterious changes, simply restoring existing functions.
    In this study, we first estimated selective effects of mutations inside mammalian protein coding sequences, under a model assuming no adaptation at the phylogenetic scale.
    We then estimated the proportion of beneficial mutations that are not adaptive innovations among all beneficial mutations at the population scale.
    Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated.
    In modern humans, it results in around 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.
\end{abstract}

@@

Adaptation is considered one of the main processes shaping the diversity of forms and functions across the tree of life\cite{darwin_origin_1859}.
It enables species to access new ecological niches and also respond to changes in their environment\cite{gavrilets_adaptive_2009}.
In evolution, adaptation is tightly linked to the notion of change at the level of the environment and to species responding to this change\cite{merrell_adaptive_1994}.
The process of adaptation leaves marks of accelerated evolution in genomes since mutations that are beneficial under new conditions fix faster than neutral mutations (fig.~\ref{fig:fitness-landscape}A).
The signature of an accelerated rate of substitutions can thus be interpreted as a sign of adaptation\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
The current availability of large-scale genomic data and the development of theoretical models have allowed to detect and quantify acceleration of substitution rates across genes and lineages, and these analyses have become common practice in evolutionary biology\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
Although these approaches have led to a better understanding of the processes determining the rates of molecular evolution, they have had the collateral effect of associating beneficial mutations with adaptive evolution.
We argue that we should dissociate beneficial mutations from adaptive mutations since adaptive evolution is not the only process that can lead to beneficial mutations\cite{charlesworth_other_2007, mustonen_fitness_2009}, and in turn, limit the use of adaptive mutations to those that are associated with adaptation to environmental change as such.
We here suggest that many beneficial mutations are not adaptive, but are rather restoring ancestral states of higher fitness.
We used large-scale genomic data to integrate analyses at both the phylogenetic and population scales to differentiate between truly adaptive mutations from mutations restoring the ancestral fitness.
Our results show that between 10 and 25\% of beneficial mutations in mammalian populations are not driven by adaptive innovations, but rather compensate the effect of ancestral deleterious mutations.

In a constant environment, a deleterious mutation can reach fixation in the population by genetic drift\cite{Ohta1992}.
Subsequently, a new mutation that is restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment has occurred\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
The restoration of the ancestral fitness can happen either at a different locus than the initial mutation, in which case it is referred to as a compensatory mutation\cite{hartl_compensatory_1996, mustonen_fitness_2009}, or at the same locus than the initial mutation, referred to as a beneficial back-mutation\cite{piganeau_estimating_2003, charlesworth_other_2007}.
While compensatory mutations induce genetic changes at the sequence level, and thus genetic diversification, beneficial back-mutations reduce the genetic diversity and do not contribute to genetic innovation.
In this study, we focus on beneficial back-mutations, which merely cancel the effect of the previous deleterious changes.
Altogether, we argue that not all beneficial mutations contribute to genetic innovation, and thus that the extent of adaptive evolution can only be correctly estimated if we can account for the number of beneficial back-mutations\cite{keightley_what_2010, rice_evolutionarily_2015}.

Given a beneficial mutation, whether it is a new innovation responding to a change in environment or a beneficial back-mutation has the same consequences on its bearer\cite{charlesworth_other_2007}.
Similarly, at the level of the population, both types of mutation will result in a positive transmission bias of the beneficial allele.
However, at the macro-evolutionary scale the consequences of these two types of mutations are fundamentally different.
While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and may help preserve well established biological systems (fig.~\ref{fig:fitness-landscape}D).
Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and, hence, in the underlying fitness landscape\cite{bazykin_changing_2015}.
On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known because we will expect to see changes from non-optimal amino acids to optimal ones.
If we wish to quantify the amount of beneficial back-mutations that have happened along a phylogeny we need to identify the selection coefficients of mutations and to achieve this we require the fitness landscape to be stable and be known.

The placental mammals represent an excellent study system to perform this analysis.
Having originated $\sim$100 Mya\cite{kumar_timetree_2017}, they diversified quickly.
Additionally, we have polymorphism data available for many species\cite{howe_ensembl_2021} and high quality protein-coding DNA alignments across the genome\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
By restricting our analysis to protein-coding orthologous genes to all the phylogeny, and further excluding any genes for which there have been claims of pervasive directional selection, we can not only assume a constant fitness landscape for each of them throughout placental mammal evolution, but actually reconstruct the fitness landscape from variable positions within the sequence along the phylogeny.
For each gene, we used a codon model of mutation-selection\cite{halpern_evolutionary_1998, mccandlish_modeling_2014} to obtain relative fitnesses for all amino-acids for each site of the sequence by fitting the model to the multi-species sequence alignment (fig~\ref{fig:method}A) and assuming that the underlying fitness landscape is stable along the phylogenetic tree.
%Mutation-selection codon models provide a nearly neutral framework of evolution taking into account non-adaptive processes impacting codon frequency\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.
From the fitness landscape, we can calculate the scaled selection coefficient $\Sphy$ for any possible amino-acid change at each position.
Under this model, a mutation with a positive $\Sphy$ would be driving the current site toward a known fitter amino-acid and would thus be considered as a beneficial back-mutation (fig~\ref{fig:method}B), given the assumed stability of the fitness landscape.

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
    \caption{
        Panel A and B: at a given site of a protein-coding DNA sequence, the different amino-acids (x-axis) have different fitnesses (y-axis).
        Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
        The protein sequence is always lagging behind the moving target defined by the amino-acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
        At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
        Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino-acid.
        At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
        At the genome level, slightly deleterious mutations that have reached fixation are supposedly scattered at different loci, such that beneficial back-mutations should also be spread across the genome.
        We thus expect that the genome-wide signature of beneficial back-mutations can be detected and quantified, even though each mutation has individually a small beneficial effect on their bearer.
    }
    \label{fig:fitness-landscape}
\end{figure*}
